Notes on $\etarho$

### 2009-09-25 docs/eta_notes/eta.tex
updated

Recall that if initially $ \overline{\rhop} = 1 /
{4\pi} \int_{\Omega_H} \rhop \; d\Omega = 0,$ 
\MarginPar{this section is here temporarily, just to better understand
what's going on here.  These notes came from an early version of
paper~III} 
there is no guarantee that $\overline{\rhop} = 0$ will hold at later
time.  To see this we start with the equation for $\rho'$, 
\begin{equation}
\frac{\partial \rhop}{\partial t} = - \nablab \cdotb (\rho_0 \Ubt)
-\nablab \cdotb (\rhop w_0 \er) - \nablab \cdotb (\rhop \Ubt )
\label{eq:rhoprimeold} \enskip ,
\end{equation}
rewritten by moving one term to the left-hand-side:
\begin{equation}
\frac{\partial \rhop}{\partial t} + \nablab \cdotb (\rhop \Ubt) 
= - \nablab \cdotb (\rho_0 \Ubt)
  -\nablab \cdotb (\rhop w_0 \er) \enskip .
\end{equation}
We integrate this over a spherical shell of thickness $2h$ at radius $r_0$, 
$\Omega_H \times (r_0-h, r_0+h)$, and normalize by the integration
volume, which is $\sim 4\pi r_0^2 \, 2h$ for small $h$, to obtain:
\begin{eqnarray}
\frac{1}{4\pi r_0^2 \, 2h} \int_{r_0-h}^{r_0+h} r^2 \; dr \int_{\Omega_H} 
    \left(\frac{\partial \rhop}{\partial t} + \nabla \cdotb ( \rhop \Ubt ) \right) \; d\Omega 
&=& - \frac{1}{4\pi r_0^2 \, 2h} \int_{r_0-h}^{r_0+h} r^2 \; dr 
\int_{\Omega_H} \nabla \cdotb \left( \rho_0 \Ubt + \rhop w_0 \er \right) \; d\Omega \nonumber  \\
&=& \left .  - \frac{1}{4\pi r_0^2 \, 2h}  \int_{\Omega_H} 
\left [ \rho_0  (\Ubt \cdot \er)  + \rhop w_0 \right] r^2
  \; d\Omega  \; \right  |_{r_0-h}^{r_0+h} \nonumber  \\
&=& 0 \nonumber \enskip ,
\end{eqnarray} 
where we have used the divergence theorem in spherical coordinates to transform
the volume integrals into area integrals over $\Omega_H.$
We see that both of the terms on the right hand side disappear. The first,
$\int_{\Omega_H} \rho_0 (\Ubt \cdot \er) d\Omega = 0$ follows from the definition of $\Ubt,$
and the second, $\int_{\Omega_H} \rhop w_0 d\Omega = w_0 \int_{\Omega_H} \rhop d\Omega = 0$
follows from our initial assumption.  
%\begin{eqnarray}
%\frac{1}{4\pi r_0^2 \, 2h} \frac{\partial}{\partial t} 
%  \int_{r_0-h}^{r_0+h} \int_{\Omega_H}  \rhop \; r^2 dr \; d\Omega &=&  
%\frac{1}{4\pi r_0^2 \, 2h} \int_{r_0-h}^{r_0+h} \int_{\Omega_H} \nabla \cdotb \left( \rho_0 \Ubt + \rhop w_0 \er \right)
%\; r^2 dr \; d\Omega &=&
%\\
%&=& \left .  - \frac{1}{4\pi r_0^2 \, 2h} \int_{\Omega_H} 
% \rhop (\Ubt \cdot \er)  \; r^2 d\Omega  \right  |_{r_0-h}^{r_0+h}
%\enskip ,
%\end{eqnarray} 
Now expanding the remaining terms and taking the limit as $h \rightarrow 0,$ we can write 
\begin{eqnarray} 
0 &=& \lim_{h\rightarrow 0} \frac{1}{4\pi r_0^2 \, 2h} \int_{r_0-h}^{r_0+h} r^2 \; dr \int_{\Omega_H} 
    \left(\frac{\partial \rhop}{\partial t} + \nabla \cdotb ( \rhop \Ubt ) \right) \; d\Omega \nonumber \\
&=& \frac{\partial}{\partial t} \left( \lim_{h\rightarrow 0} \frac{1}{4\pi r_0^2} \frac{1}{2h} 
  \int_{r_0-h}^{r_0+h} r^2 \; dr \int_{\Omega_H}  \rhop \;d\Omega \right) +
\lim_{h\rightarrow 0} \left( \frac{1}{4\pi r_0^2 \, 2h}  
  \int_{r_0-h}^{r_0+h} r^2 \; dr \int_{\Omega_H}  \nabla \cdotb ( \rhop \Ubt ) \; d\Omega \right) \nonumber \\
&=& \frac{\partial}{\partial t} \left(\frac{1}{4\pi} \int_{\Omega_H}  \rhop \; d\Omega \right) +
\lim_{h\rightarrow 0} \left( \left .  \frac{1}{4\pi r_0^2 \, 2h}  
\left [ \int_{\Omega_H} \rhop (\Ubt \cdot \er) \; d\Omega  \right]  r^2\; \right |_{r_0-h}^{r_0+h} \right) \nonumber  \\
&=&  \frac{\partial}{\partial t} \overline{\rhop} + \lim_{h\rightarrow 0} \left( \left .  \frac{1}{r_0^2 \, 2h}
\overline{[\rhop (\Ubt \cdot \er)]} \; r^2 \; \right |_{r_0-h}^{r_0+h} \right) \nonumber  \\
&=&  \frac{\partial}{\partial t} \overline{\rhop} + 
\lim_{h\rightarrow 0} \frac{1}{r_0^2 \, 2h} \int_{r_0-h}^{r_0+h} \nabla \cdot \left[\overline{\left(\rhop\Ubt \cdot \er\right)} \er \right]\;  r^2 \; dr \nonumber \\
&=&  \frac{\partial}{\partial t} \overline{\rhop} + \nabla \cdot \left[ \overline{\left(\rhop \Ubt \cdot \er\right)} \er \right]
\end{eqnarray} 
again using the divergence theorem, extracting the time derivative from the spatial integral,
and switching the order of operations as appropriate.

In short, 
\begin{equation}
\frac{\partial}{\partial t} \overline{\rhop} = 
- \nabla \cdot \left [ \overline{\left(\rhop \Ubt \cdot \er\right)} \, \er \right ] = 
- \nabla \cdot ( \etarho \er )
 \enskip .\label{eq:rhopbar} 
\end{equation}

>>>>

In Paper III, we introduced a mixing term, $\etarho$, to the base state density
evolution equation (see eq. [29] of Paper III), with the objective of
keeping the base state density equal to the average of the density
over a layer.  Here we derive $\etarho$ for spherical problems.
For this derivation, it is best to define the average in terms of spherical coordinates,
\begin{equation}
\overline{\phi(r)} = \frac{1}{4\pi} \int_{\Omega_H} \phi(r,\theta,\phi) d\Omega
\end{equation}
where $\int_{\Omega_H} d\Omega = 4\pi$ represents the integral over
the angles $\theta$ and $\phi$ at constant radius.

Recall from Paper III that if initially $\overline{\rho'} = 0$,
there is no guarantee that $\overline{\rho'} = 0$ will hold at later
time.  To see this, recall equation (\ref{eq:Perturbational Density}),
written in a slightly different form:
\begin{equation}
\frac{\partial\rho'}{\partial t} + \nabla\cdot\left(\rho'\Ub\right) = -\nabla\cdot\left(\rho_0\Ubt\right).
\end{equation}
We integrate this over a spherical shell of thickness $2h$ at radius $r_0$, i.e.,
$\Omega_H \times (r_0-h, r_0+h)$, and normalize by the integration
volume, which is $\sim 4\pi r_0^2  2h$ for small $h$, to obtain:
\begin{eqnarray}
\frac{1}{4\pi r_0^2 2h}\int_{r_0-h}^{r_0+h}r^2 dr\int_{\Omega_H}\left[\frac{\partial\rho'}{\partial t} + \nabla\cdot\left(\rho'\Ub\right)\right]d\Omega 
&=& - \frac{1}{4\pi r_0^2 2h}\int_{r_0-h}^{r_0+h}r^2 dr\int_{\Omega_H}\nabla\cdot\left(\rho_0\Ubt\right)d\Omega \nonumber \\
&=& \left. -\frac{1}{4\pi r_0^2 2h}\int_{\Omega_H}\left[\rho_0\left(\Ubt\cdot\eb_r\right)\right]r^2 d\Omega\right|_{r_0-h}^{r_0+h} \nonumber \\
&=& 0,
\end{eqnarray} 
where we have used the divergence theorem in spherical coordinates to transform
the volume integral on the right hand side into an area integral over $\Omega_H$.  
We see that the right hand side disappears since 
$\int_{\Omega_H}\rho_0(\Ubt\cdot\eb_r)d\Omega=0$,  which follows from the definition 
of $\Ubt$.  Now, expanding the remaining terms and taking the limit as 
$h\rightarrow 0,$ we can write
\begin{eqnarray} 
0 &=& \lim_{h\rightarrow 0} \frac{1}{4\pi r_0^2 2h} \int_{r_0-h}^{r_0+h} r^2  dr \int_{\Omega_H} \left[\frac{\partial \rho'}{\partial t} + \nabla\cdot(\rho'\Ub)\right] d\Omega \nonumber \\
&=& \frac{\partial}{\partial t} \left( \lim_{h\rightarrow 0} \frac{1}{4\pi r_0^2} \frac{1}{2h} \int_{r_0-h}^{r_0+h} r^2  dr \int_{\Omega_H}  \rho' d\Omega \right) + \lim_{h\rightarrow 0} \left[\frac{1}{4\pi r_0^2  2h} \int_{r_0-h}^{r_0+h} r^2  dr \int_{\Omega_H}  \nabla \cdot ( \rho' \Ub )  d\Omega\right] \nonumber \\
&=& \frac{\partial}{\partial t} \left(\frac{1}{4\pi} \int_{\Omega_H}  \rho'  d\Omega \right) + \lim_{h\rightarrow 0} \left\{\left.\frac{1}{r_0^2 2h}\left[\frac{1}{4\pi}\int_{\Omega_H} \rho' (\Ub \cdot \eb_r)  d\Omega  \right]  r^2 \right |_{r_0-h}^{r_0+h} \right\} \nonumber  \\
&=&  \frac{\partial}{\partial t} \overline{\rho'} + \lim_{h\rightarrow 0} \left\{ \left .  \frac{1}{r_0^2  2h} \left[ \overline{\rho' (\Ub \cdot \eb_r)} \right]  r^2  \right |_{r_0-h}^{r_0+h} \right\} \nonumber  \\
&=&  \frac{\partial}{\partial t} \overline{\rho'} + \lim_{h\rightarrow 0} \frac{1}{r_0^2 2h} \int_{r_0-h}^{r_0+h} \nabla \cdot \left[\overline{\rho' (\Ub \cdot \eb_r)} \eb_r \right]  r^2  dr \nonumber \\
&=&  \frac{\partial}{\partial t} \overline{\rho'} + \nabla \cdot \left[ \overline{\rho' (\Ub \cdot \eb_r)} \eb_r \right]
\end{eqnarray} 
again using the divergence theorem, extracting the time derivative from the spatial integral,
and switching the order of operations as appropriate.

In short, 
\begin{equation}
\frac{\partial}{\partial t} \overline{\rho'} = - \nabla\cdot\left[\overline{\rho'\left(\Ub\cdot\eb_r\right)}\eb_r\right] = -\nabla\cdot\left(\etarho\eb_r\right),
\end{equation} 
and thus, $\etarho = \overline{\left(\rho'\Ub\cdot\eb_r\right)}$.
